A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-05-01 paper/paper.tex
paper: major restructuring

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds83} interprets types as relations characterizing the observational behavior of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has the structure of a cubical set~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates or families instead of relations or graphs, in which case, we obtain a form of realizability~\cite{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}.

@@

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}

In this section, we give a uniform definition of semi-simplicial and semi-cubical sets which we believe is folklore. The uniformity allows the definition of a notion of $\nu$-sets that subsumes semi-simplicial and semi-cubical sets.

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner into a category which we call $\nu$-semi-shape category. The morphisms of the $\nu$-semi-shape category are sequences of elements of a set $\nu$ of arbitrary cardinal, extended with $\kstar$, so that the following holds.

\section{Indexed representation and parametricity\label{sec:indexed-param}}
In this section, we introduce some preliminaries on interpreting sets in type theory, followed by a small section on the indexed presentation, before relating it to parametricity in \ref{sec:rel-param}.

\subsection{Preliminaries}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~\cite{agda23}, Coq~\cite{coq22} or Lean~\cite{lean15}. In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

@@

\subsection{Fibered versus indexed representation}
There are different ways to represent a family of sets, commonly known as the fibered and indexed representations. In type theory, they are equivalent, and this equivalence can be formulated as:
\begin{equation*}
  (\Sigma S: \U. (S \rightarrow T)) \simeq (T \rightarrow \U)
\end{equation*}

@@

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \rightarrow X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions on the functor (e.g. that $\delta^R \circ \delta^{R\kstar} = \delta^{(R\kstar)\circ(R)} = \delta^{RR} = \delta^{(\kstar R)\circ(R)} = \delta^R \circ \delta^{\kstar R}$, etc.). Based on these coherence conditions, the types of $X_0, X_1, X_2$ happen to have the form:
\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}

\subsection{Relating to parametricity\label{sec:rel-param}}
The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
   & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq                                                      \\
   & \Pi a_L: A.\, \Pi a_R: A.\, \Pi a_\kstar: A_\kstar(a_L)(a_R).\, B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

@@

\subsection{Formal construction in ITT\label{sec:itt}}
\renewcommandx{\cohf}{\coh{frame}[][][][][][][][]}
\renewcommandx{\cohl}{\coh{layer}[][][][][][][][]}

@@

\subsection{Details on the mechanization\label{sec:mech}}
\renewcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\renewcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\renewcommandx{\painting}[2][1,2]{\prim{painting}[][#1][#2][][]}
\renewcommandx{\restrf}[4][1,2,3,4]{\restr{frame}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrl}[4][1,2,3,4]{\restr{layer}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrc}[4][1,2,3,4]{\restr{painting}[][#3][#4][#1][#2][][]}
\renewcommandx{\cohf}[1][1]{\coh{frame}[][][][][][#1][][]}
\renewcommandx{\cohl}[1][1]{\coh{layer}[][][][][][#1][][]}
\renewcommandx{\cohc}[1][1]{\coh{painting}[][][][][][#1][][]}

>>>>

\section{Introduction}
% - semisimplicial/semicubical sets in indexed form
%   - clarify indexed/fibered, justify advantages of the respective choice
%     - the presentation matters when expressed in indexed way!
%   - compare to indexed VV, previous indexed HH, indexed Pang-Luo
% - iterated unary/binary parametricity has a semi-simplicial/semi-cubical structure
% - nu-types as a uniform definition of semi-simplicial/semi-cubical
%   iterated parametricity gives an alternative presentation of indexed semi-simplicial/semi-cubical
%   this is what we described in the paper
% - perspective: models of cubical type theory using indexed cubical sets, hence preserving more definitial rules
% - mention also the semi-simplicial types problem

In the context of functional programming, Reynolds' parametricity~\cite{reynolds83} interprets types as relations characterizing the observational behavior of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has the structure of a cubical set~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates or families instead of relations or graphs, in which case, we obtain a form of realizability~\cite{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}.

@@

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}
In this section, we generalize semi-simplicial and semi-cubical sets to, what we call, $\nu$-sets, subsuming the earlier defintions. However, we also explicitly work out the cases of semi-simplical and semi-cubical sets, in a way that is similar to a previous art~\cite{aitchison86}.

\subsection{\texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner into a category which we call $\nu$-semi-shape category. The morphisms of the $\nu$-semi-shape category are sequences of elements of a set $\nu$ of arbitrary cardinal, extended with $\kstar$, so that the following holds.

@@

\section{Preliminaries}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~\cite{agda23}, Coq~\cite{coq22} or Lean~\cite{lean15}. In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

@@

\section{Comparison of fibered and indexed presentations}
%     - standard augmented triangle:
%       Inductive points := A | B | C.
%       Inductive lines := AB | BC | AC.
%       Inductive surface := ABC.

%       {X n := match n with 0 => unit | 1 => points | 2 => lines | 4 => surface | n => empty_set;
%        d {n} i (Hi:i<n) (x:X n) := match n, i, x with
%        | 0, _, _ => assert false
%        | 1, _, _ => tt
%        | 2, 0, AB => B
%        | 2, 1, AB => A
%        | 2, 0, BC => C
%        | 2, 1, BC => B
%        | 2, 0, AC => C
%        | 2, 1, AC => A
%        | 3, 0, ABC => BC
%        | 3, 1, ABC => AC
%        | 3, 2, ABC => AB
%        | n.+4, i, _ => assert false
%        end;
%        alpha n i j (Hi:i<n) (Hj:j<i) (x : X n) : d i Hi (d j _ x) = d (j-1) _ (d i _ x)
%        := match n, i, j, x with
%        | 0, _, _, _ => assert false
%        | 1, _, _, _ => assert false
%        | 2, _, _, _ => proof_irrelevance_of_unit : tt = tt
%        | 3, 1, 0, ABC => proof_irrelevance_in_points : C = C
%        | 3, 2, 0, ABC => proof_irrelevance_in_points : B = B
%        | 3, 2, 1, ABC => proof_irrelevance_in_points : A = A
%        | n.+4, i, _ => assert false
%        end
%        }
%        {hd tt := unit;
%         tl := {hd _ (* (tt;tt) *) := points; (* 3 points *)
%                tl := {hd _ (* ((tt;(tt;a));b) *) := unit; (* unique line over any a and any b *)
%                       tl := {hd _ (* ((tt;(tt;(a;(b;tt))));((c;tt);tt)) *) := unit; (* unique triangle over ab, bc, and ac *)
%                              tl := cofix s := {hd _ := empty_set; tl := s}}}}}
%        {hd _ := unit; tl :=
%        {hd _ := points; tl :=
%        {hd _ := unit; tl :=
%        {hd _ := unit; tl := cofix s := {hd _ := empty_set; tl := s}}}}} : AugmentedSemiSimplicial

%        standard n : AugmentedSemiSimplicial
%        := {hd _ := unit; tl :=
%           {hd _ := [0;n]; tl :=
%            unit unit unit ... n times ... unit; empty_set; ... }}
There are different ways to represent a family of sets, commonly known as the fibered and indexed representations. In type theory, they are equivalent, and this equivalence can be formulated as:
\begin{equation*}
  (\Sigma S: \U. (S \rightarrow T)) \simeq (T \rightarrow \U)
\end{equation*}

@@

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \rightarrow X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions on the functor (e.g. that $\delta^R \circ \delta^{R\kstar} = \delta^{(R\kstar)\circ(R)} = \delta^{RR} = \delta^{(\kstar R)\circ(R)} = \delta^R \circ \delta^{\kstar R}$, etc.). Based on these coherence conditions, the types of $X_0, X_1, X_2$ happen to have the form:
\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}
\section{Relating to parametricity\label{sec:rel-param}}
The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
   & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq                                                      \\
   & \Pi a_L: A.\, \Pi a_R: A.\, \Pi a_\kstar: A_\kstar(a_L)(a_R).\, B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

@@

\section{Details on the mechanization\label{sec:mech}}
\subsection{Formal construction in ITT\label{sec:itt}}
\renewcommandx{\cohf}{\coh{frame}[][][][][][][][]}
\renewcommandx{\cohl}{\coh{layer}[][][][][][][][]}

@@

\subsection{Finer details}
\renewcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\renewcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\renewcommandx{\painting}[2][1,2]{\prim{painting}[][#1][#2][][]}
\renewcommandx{\restrf}[4][1,2,3,4]{\restr{frame}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrl}[4][1,2,3,4]{\restr{layer}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrc}[4][1,2,3,4]{\restr{painting}[][#3][#4][#1][#2][][]}
\renewcommandx{\cohf}[1][1]{\coh{frame}[][][][][][#1][][]}
\renewcommandx{\cohl}[1][1]{\coh{layer}[][][][][][#1][][]}
\renewcommandx{\cohc}[1][1]{\coh{painting}[][][][][][#1][][]}
